Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
paper full
update conclusions

Setting $\Edist = \binom{n}{2}$ will generate random permutations at early stages, slowing the algorithm convergence. 
A purely explotation approach will set $\Edist = 0$  generating $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance   controls the exploration-explotation tradeoff as a function of the number of function evaluations ($m$) that the algorithm performs, thus \manuel{it follows that}{} the behaviour is different for different choices of $m$. Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}.

@@

\subsubsection{Computing environment}
% Due to the computational effort required, we used several computing systems to
% run experiments.
% All UMM experiments were either on run on Intel Xeon
% \emph{Broadwell} E5-2683v4 CPUs at 2.10\,GHz or Intel Xeon \emph{Haswell} E5-2680v3
% CPUs at 2.50\,GHz, both of them with 32\,GB RAM. 
 %
Experiments were run on Intel Xeon \emph{Ivybridge} E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM running CentOS Linux release 7 (Core).
%The operating system was CentOS Linux release 7 (Core).

@@

\paragraph{Convergence analysis of CEGO}
The red and the gold lines in the plots shown in Fig.~\ref{fig:results}
correspond to the two variants of CEGO the original \CEGOorig and our modified
version \CEGOinv. The different between these two variants is overwhelmingly in
favour of \CEGOinv. With the exception of the \texttt{reC19} PFSP instance, the
search behavior of \CEGOorig resembles a random search, after a brief
improvement phase at the start of the run. By contrast, \CEGOinv shows a very
fast improvement in fitness within the first 50 evaluations that tends to
converge around the 150$^\text{th}$ evaluation, showing little improvement
after that. Since the only difference between these two variants is the
transformation from rankings to orderings before evaluating the objective
function (Sec.~\ref{sec:exper_ro}), we must conclude that this difference has
tremendous effect in performance. Therefore, we will focus our analysis on
\CEGOinv.

@@

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-04} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

@@

\section{Conclusions}\label{sec:conclusions}
In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (here, up to $100$, $200$ and $400$ evaluations).
Our experiments presented here show that UMM is able to match the results of
the state-of-the-art Bayesian optimizer for expensive black-box
optimization if the computation time of the optimizers is not a concern.

In addition, UMM is specially well-suited for budget-limited scenarios that are
still time-sensitive, e.g., when the overhead incurred by the optimizer should
be no more than a few minutes added to each fitness evaluation. By comparison,
CEGO may require up to one hour per fitness evaluation, with the time
increasing non-linearly with permutation size and larger budgets.  Hence, UMM
is a computationally feasible alternative for relatively large problem sizes.

>>>>

Setting $\Edist = \binom{n}{2}$ will generate random permutations at early stages, slowing the algorithm convergence. 
A purely explotation approach will set $\Edist = 0$  generating $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance   controls the exploration-explotation tradeoff as a function of the number of function evaluations ($m$) that the algorithm performs, thus the behaviour is different for different choices of $m$. Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}.

@@

\myparagraph{Computing environment}
% Due to the computational effort required, we used several computing systems to
% run experiments.
% All UMM experiments were either on run on Intel Xeon
% \emph{Broadwell} E5-2683v4 CPUs at 2.10\,GHz or Intel Xeon \emph{Haswell} E5-2680v3
% CPUs at 2.50\,GHz, both of them with 32\,GB RAM. 
 %
Experiments were run on Intel Xeon \emph{Ivybridge} E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM running CentOS Linux release 7 (Core).
%The operating system was CentOS Linux release 7 (Core).

@@

\myparagraph{Convergence analysis of CEGO}
The red and the gold lines in the plots shown in Fig.~\ref{fig:results}
correspond to the two variants of CEGO the original \CEGOorig and our modified
version \CEGOinv. The different between these two variants is overwhelmingly in
favour of \CEGOinv. With the exception of the \texttt{reC19} PFSP instance, the
search behavior of \CEGOorig resembles a random search, after a brief
improvement phase at the start of the run. By contrast, \CEGOinv shows a very
fast improvement in fitness within the first 50 evaluations that tends to
converge around the 150$^\text{th}$ evaluation, showing little improvement
after that. Since the only difference between these two variants is the
transformation from rankings to orderings before evaluating the objective
function (Sec.~\ref{sec:exper_ro}), we must conclude that this difference has
tremendous effect in performance. Therefore, we will focus our analysis on
\CEGOinv.

@@

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-04} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

@@

\section{Conclusions}\label{sec:conclusions}

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
expensive black-box combinatorial problems on permutation landscapes when the
budget of fitness evaluations is severely limited (here, up to $100$, $200$ and
$400$ evaluations).  Our experiments presented here show that UMM is not far
away from, and sometimes  is able to match, the results of the state-of-the-art
Bayesian optimizer for expensive black-box optimization.
%if the computation time of the optimizers is not a concern.

In addition, UMM is specially well-suited for budget-limited scenarios that are
still time-sensitive, e.g., when the overhead incurred by the optimizer should
be no more than a few minutes added to each fitness evaluation. By comparison,
CEGO may require around one day to perform 400 fitness evaluations, with the time
increasing non-linearly with permutation size and larger budgets.  Hence, UMM
is a computationally feasible alternative for relatively large problem sizes.
